Estimating Uncertainty for Tipping Points with Deep Learning

### 2022-05-12 manuscript/manuscript.Rmd
tweaking plots, some writing
added a bit explicitly about frameworks

<!-- some more verbose bits about bifurcation models probably belongs in discussion instead -->
We will focus the analysis of several different forecasting scenarios based around two classic models in population ecology: 
Robert May's consumer-resource model [@May1979], and the Nicholson-Bailey parasatoid-host model [@Nicholson1935]. 
While 'tactical' population dynamics models used to inform decision making in fields such as fisheries are often much more complex, these simple models nevertheless exhibit rich nonlinear dynamics thought to be typical in ecological systems [@]. <!-- MFL: Using 'tactical' might throw readers off that aren't familiar with tact vs strat distinction -->
These textbook models have been well studied and form the basis of half a century of research in ecology, including much recent work on topics such as resilience and tipping points [] which has had important theoretical and practical management outcomes [@Folke2009; @].
May's model exhibits alternative stable states. 
In this one-dimensional model, transitions between these states can occur due to intrinsic stochasticity, external forcing, or the gradual environmental change that results in a catastrophic saddle-node bifurcation and generates hysteresis. 
The Nicholson-Bailey model is a two species model which contains a Hopf bifurcation, a non-catastrophic bifurcation which either creates or destroys a limit cycle -- a stable oscillatory pattern.

Assessing the accuracy of forecasting methods in the face of such bifurcation dynamics is particularly important question for ecological systems and global environmental change problems.
Bifurcations represent the kind of non-linear responses complex systems can make as the result of slowly changing parameters (i.e. the 'environment' in which those dynamics take place.)
This can create a particularly challenging forecasting task when such transitions have not been previously observed in the same system, requiring the forecast to anticipate dynamics for which there are no corresponding analogue in the historical data to date. 
No-analog scenarios [@Williams]

@@

If the environment slowly alters one of the parameters (say, the encounter efficiency, `h_t`, in our formulation), one of the stable nodes moves closer and closer to the unstable saddle point, leading to a bifurcation that destroys the stable state, leaving the system to suddenly transition to the alternative stable state.
Saddle-node bifurcations (also known as fold bifurcations) also create a phenomenon known as hysteresis, where it is not sufficient to restore the environment to the previous parameter values to recover the previous state. 
Unlike the Hopf bifurcation which exhibits a continuous transition from a stable node to a small limit cycle that then grows, the saddle-node transition is a discontinuous or so-called 'catastrophic' bifurcation.
Due both to this sudden, catastrophic nature of the transition and the difficulty in reversing the shift after it has occurred, saddle-node bifurcations have been the subject of intense study.
These dynamics have long been identified as an acute challenge to forecasting, with effort focused primarily on identifying any 'early warning signs' that a catastrophic bifurcation might occur at all [@Scheffer2009; @Scheffer2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs. <!-- MFL: I think we can make the first half of this sentence clearer. I think some readers might get lost here. -->
Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dei2011; @Dei2012] to whole-ecosystem experiments [@Carpenter2011], and postulated as a model for global change [@Barnosky2014].

@@

To compare forecasts, we focus exclusively on metrics of forecast skill which satisfy @Gneiting2007's property of a strictly proper score.
This ensures the very desirable behavior that which no probabilistic forecast $Q(x,t)$ can have a score as high as that of the true process $P(x,t)$ on average. 
In other words, while it is possible for any of the models considered to _overfit_ the data against which they are trained, i.e. have a higher likelihood than the true process, it is not possible for these models to overfit the data against which they are scored. <!-- MFL: I think we can take out the parenthetical here and assume the reader knows what we mean by overfitting-->
It is worth noting that this property applies specifically to probabilistic forecasts, and is not true of all metrics often used to compare forecasts. <!-- MFL: hmm I am not completely following this point here -->
Concerns about over-fitting arise in most types of model estimation and are a particularly acute concern to machine learning methods due to the bias-variance trade-off [@]. <!-- MFL: I don't think I nailed the overfitting part in ML section but we probably want to reference this here-->
This makes the use of strictly proper scoring especially relevant in assessing machine learning predictions.

While all strictly proper scores have this property, this does not guarantee that all such metrics will share the same relative ranking between forecasts.
We will focus on two of the most common such skill metrics, CRPS score and logs score [e.g. see @scoringRules; @Gneiting2014].
Of the two, the logs probability score puts a much a greater penalty on unexpected observations than CRPS

@@

@@

```{r}
shift |> 
  ggplot(aes(forecasting_model, score, color=forecasting_model, fill=forecasting_model)) + 
  geom_violin() +
  facet_grid(metric ~ simulation, scales = "free") + 
  scale_y_log10() +
  theme_bw() + 
  coord_flip()

  ggplot(aes(simulation, score, color=forecasting_model, fill=forecasting_model)) + 
  geom_violin() +
  facet_wrap(~ metric, scales = "free", ncol=1) + 
  scale_y_log10() + 
  theme_bw()
```

```{r}
shift |>
  ggplot(aes(forecasting_model, score, color=forecasting_model)) + 
  geom_jitter(alpha = 0.01, show.legend = FALSE) +
  facet_grid(metric ~ simulation, scales = "free") + 
  scale_y_log10() +
  theme_bw() + 
  coord_flip()

@@

```{r}
# scores over time, averaged over replicates
over_time <- shift |> 
  group_by(t, simulation,forecasting_model, metric ) |> 
  summarise(score = mean(score))

over_time |> 
  ggplot(aes(t, score, col = forecasting_model)) + 
  geom_point(size=1) +
  facet_grid(metric ~ simulation,
             scales="free") + 
  theme_bw() + scale_y_log10()

>>>>

<!-- some more verbose bits about bifurcation models probably belongs in discussion instead -->
We will focus the analysis of several different forecasting scenarios based around two classic models in population ecology: 
Robert May's consumer-resource model [@May1979], and the Nicholson-Bailey parasatoid-host model [@Nicholson1935]. 
While population dynamics models used to inform decision making in fields such as fisheries are often much more complex, these simple models nevertheless exhibit rich nonlinear dynamics thought to be typical in ecological systems [@]. 
These textbook models have been well studied and form the basis of half a century of research in ecology, including much recent work on topics such as resilience and tipping points [] which has had important theoretical and practical management outcomes [@Folke2009; @].
May's model exhibits alternative stable states. 
In this one-dimensional model, transitions between these states can occur due to intrinsic stochasticity, external forcing, or the gradual environmental change that results in a catastrophic saddle-node bifurcation and generates hysteresis. 
The Nicholson-Bailey model is a two species model which contains a Hopf bifurcation, a non-catastrophic bifurcation which either creates or destroys a limit cycle -- a stable oscillatory pattern.

Assessing the accuracy of forecasting methods in the face of such bifurcation dynamics is particularly important question for ecological systems and global environmental change problems.
Bifurcations represent the kind of non-linear responses complex systems can make as the result of slowly changing parameters (i.e. the 'environment' in which those dynamics take place.)
This can create a particularly challenging forecasting task when such transitions have not been previously observed in the same system, requiring the forecast to anticipate dynamics for which there are no corresponding analogue in the historical data to date.

No-analog scenarios [@Williams]

We provide fully reproducible coded examples in R and python for performing, scoring, and visualizing each of the forecasts considered here.
After considerable time considering alternative frameworks, we have emphasized those which best met our requirements for performance, ease-of-use, flexibility, and support for the latest probabilistic machine learning models for forecasting. 
Most of our forecasts use the `darts` framework, a sophisticated and well documented python library with support for a wide range of methods.
Our model-based MCMC forecasts use the `greta` framework, an R library that uses python-based `tensorflow` probability to achieve better performance.
While python-based frameworks currently have the edge in performance and access modern ML algorithms, they lag behind in attention to statistical issues such as the computation of strictly proper skill scores (discussed below.) 
Thus our examples of scoring and visualization will rely on a collection of R packages, in particular, `scoringRules` for the efficient calculation of Continuous Ranked Probability Score (CRPS) and logarithmic probability scores for forecast ensembles [@scoringRules].
We expect greater convergence between methods available in R and python in the future, as already illustrated in the example of `greta`.
Complete code for all examples presented here can be found at https://github.com/boettiger-lab/mee_tipping_point_forecasting.

@@

If the environment slowly alters one of the parameters (say, the encounter efficiency, `h_t`, in our formulation), one of the stable nodes moves closer and closer to the unstable saddle point, leading to a bifurcation that destroys the stable state, leaving the system to suddenly transition to the alternative stable state.
Saddle-node bifurcations (also known as fold bifurcations) also create a phenomenon known as hysteresis, where it is not sufficient to restore the environment to the previous parameter values to recover the previous state. 
Unlike the Hopf bifurcation which exhibits a continuous transition from a stable node to a small limit cycle that then grows, the saddle-node transition is a discontinuous or so-called 'catastrophic' bifurcation.
Due both to this sudden, catastrophic nature of the transition and the difficulty in reversing the shift after it has occurred, saddle-node bifurcations have been the subject of intense study.
Tipping point dynamics have long been identified as an important but difficult challenge for forecasting [e.g. @Scheffer2001]. 
Much effort in the ecological literature so far has focused on identifying any 'early warning signs' that a catastrophic bifurcation might occur at all [@Scheffer2009; @Scheffer2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
Tipping points resulting from Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dei2011; @Dei2012] to whole-ecosystem experiments [@Carpenter2011], and postulated as a model for global change [@Barnosky2014].

@@

To compare forecasts, we focus exclusively on metrics of forecast skill which satisfy @Gneiting2007's property of a strictly proper score.
This ensures the very desirable behavior that which no probabilistic forecast $Q(x,t)$ can have a score as high as that of the true process $P(x,t)$ on average. 
In other words, while it is possible for any of the models considered to _overfit_ the data against which they are trained, i.e. have a higher likelihood than the true process, it is not possible for these models to overfit the data against which they are scored.
It is worth noting that this property applies specifically to probabilistic forecasts and not point forecasts. Not all common metrics metrics often used to compare forecasts are strictly proper -- such as the average root-mean-square error or the average absolute error. <!-- MFL: hmm I am not completely following this point here. CB: added some examples of non-strictly-proper scoring metrics. For instance, {darts} has a whole 'metrics' class, which does not include any strictly proper metrics (because ML researchers don't care about statistical theorems?) -->
Concerns about over-fitting arise in most types of model estimation and are a particularly acute concern to machine learning methods due to the bias-variance trade-off [@Mehta2019]. <!-- MFL: I don't think I nailed the overfitting part in ML section but we probably want to reference this here-->
This makes the use of strictly proper scoring especially relevant in assessing machine learning predictions.

Not even all strictly proper scores will agree on the same relative ranking between forecasts.
We will focus on two of the most common such skill metrics, CRPS score and logs score [e.g. see @scoringRules; @Gneiting2014].
Of the two, the logs probability score puts a much a greater penalty on unexpected observations than CRPS, and may be more suitable when the occurrence of unexpected events incurs a particularly high cost.

@@

We examine forecast skill for each of the six forecasting methods (MCMC, ARIMA, block-RNN, GRU, LSTM, and transformer) in each of our three scenarios (Hopf bifucation, saddle-node bifurcation, stochastic transition).
Using model-based simulations allows us to examine performance against an ensemble of replicates of the "true" process, which further helps identify differences that may occur solely due to chance.

@@

```{r}
shift |>
  ggplot(aes(forecasting_model, score, color=forecasting_model)) + 
  geom_jitter(alpha = 0.006, show.legend = FALSE) + 
  coord_flip() +
  facet_grid(simulation ~ metric, scales = "free") + 
  scale_y_log10() +
  theme_bw()

@@

```{r}
# scores over time, averaged over replicates
over_time <- shift |> 
  group_by(t, simulation,forecasting_model, metric ) |> 
  summarise(sd = sd(score),
            score = mean(score),
            .groups = "drop") |>
  group_by(simulation,forecasting_model, metric) |> 
  mutate(t = t - min(t)) |> ungroup()

p1 <- over_time |> filter(metric == "crps") |>
  ggplot(aes(t, score, col = forecasting_model)) + 
  geom_pointrange(aes(y = score, ymin = score - sd, ymax = score + sd),
                  shape=".", alpha=0.4, show.legend=FALSE) +
  geom_line(show.legend=FALSE) +
  facet_wrap(~simulation, scales="free", ncol=1) + 
  theme_bw()

p2 <- over_time |> filter(metric == "logs") |>
  ggplot(aes(t, score, col = forecasting_model)) + 
  geom_pointrange(aes(y = score, ymin = score - sd, ymax = score + sd),
                  shape=".", alpha=0.4, show.legend=FALSE) +
  geom_line(show.legend=FALSE) +
  facet_wrap(~simulation, scales="free", ncol=1) + 
  scale_y_log10() +
  theme_bw()

p1 + p2
```
```{r}
over_time |> filter(metric == "logs") |>
  ggplot(aes(t, score, col = forecasting_model)) + 
  geom_pointrange(aes(y = score, ymin = score - sd, ymax = score + sd),
                  shape=".", alpha=0.4, show.legend=FALSE) +
  geom_line(show.legend=FALSE) +
 # facet_wrap(~simulation, scales="free", ncol=1) + 
  facet_grid(simulation ~ forecasting_model, scales="free") + 
  theme_bw() + scale_y_log10()
```
